% Write acknowledgement here as LaTeX text

Foremost, Thanks to Almighty for whose given strength makes us able to complete our research successfully.  


After that, we would like to thank our honorable supervisor Dr. Md. Maruf Hassan, Associate Professor, Department of Computer Science Engineering in Southeast University. This research behind it would not have been possible without the exceptional support of our supervisor. His enthusiasm, knowledge, and exacting attention to detail have been an inspiration and kept us on track from our first encounter with machine learning to the final draft of this paper.


We would also like to thank Mr. Shahriar Manzoor, Associate Professor \& Chairman, Department of Computer Science and Engineering, Southeast University for making Dr. Md. Maruf Hassan, our supervisor. We also want to thank our whole faculties of the department of Computer Science and Engineering, Southeast university for their encouragement and motivation. Finally, it is true with pleasure and appreciation that we acknowledge our contributions and day after day hard work with proper responsibility. 


In addition, we owe a sincere debt of gratitude to each and every one of our great group members for their thought-provoking debates, unwavering commitment, and endless sleepless hours spent together, which were essential to finishing this project ahead of schedule. We were able to finish this report within the allotted time thanks to their unflagging encouragement and assistance. 


Finally, we want to say that all of those kindnesses were indispensable to complete our research.
